### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix subst. Fix typo.
IDesc@ICFP: no 'beauty of the exercise'

This concludes our levitation work. Beyond the beauty of the exercise,
there are several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
    \pair{\ListCons{\DVar{}}{E}}
         {\pair{\DSigma{X}{\Unit}}{D}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

@@

\note{ Triple check. That doesn't work, right? }

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
           \Mu{(\FreeMonad{D}{Y})}                            \\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: \DVar{}\: X\: Y\: \sigma\: x  & \sigma x     \\
\apply\: xs\:      X\: Y\: \sigma\: ys & ys
\end{array}
}\]

>>>>

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

\[\stk{
\FreeMonad{\_} : \TagDesc \To \Set \To \TagDesc \\
\FreeMonad{\pair{E}{D}{}}\:X \mapsto
    \pair{\ListCons{\DVar{}}{E}}
         {\pair{\DSigma{X}{\DUnit}}{D}{}}{}
}\]

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\DUnit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
           \Mu{(\FreeMonad{D}{Y})}                            \\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}     & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]
